---
title: "Coping Methods App Documentation"
author: "Pooja Rajendran Raju"
date: "Aug 5, 2024"
quarto-required: ">=1.3.0"
output:
    html: 
        code-fold: true
        toc: true
        toc-depth: 2
categories: [App]
---


# Coping Methods App

The **Coping Methods App** is an interactive Shiny application included with this package.
It enables users to explore mental-health–related data through visualisations such as:

- depression trends across age groups  
- year-to-year changes  
- comparison of coping strategies between countries  

Because the app runs locally, this page provides static screenshots, key insights,
and clear explanations to help readers understand the functionality before launching it.

To run the app:

```{r, eval=FALSE}
library(copingmethods)
launch_copingmethods()
```

---

# 1. Depression by Age — Country View

::: {.card style="padding: 1.2rem; margin-bottom: 1.5rem;"}

###  What this visualisation shows

This tab displays **depression percentages across age groups** for a selected **country/entity** and **year**.

The screenshot below shows **Afghanistan (1990)**.  
The bar chart visualises depression levels across age groups, while the table on the right provides:

- depression (%) for each age group  
- differences relative to the comparison year  

::: {.center}
<div style="text-align:center; margin-top: 1rem; margin-bottom: 1rem;">
  <embed src="img1.pdf" type="application/pdf" width="100%" height="900px" />
</div>
:::

*Figure 1: Depression percentages by age group for Afghanistan in 1990.*

### Key insights
::: {.callout-tip}
- Depression increases steadily across age groups in the displayed dataset.  
- The table provides precise numerical values alongside the visual chart.  
- This view makes it easy to compare age-specific depression profiles across countries and years.  
:::

:::

---

# 2. Year-to-Year Comparison (Trend Analysis)

::: {.card style="padding: 1.2rem; margin-bottom: 1.5rem;"}

###  What this visualisation shows

This tab compares **how depression percentages have changed over time** relative to a baseline year.

The screenshot below shows **America (IHME GBD), 2009**, compared against **1990**.

The plot includes:

- depression (%) by age group in 2009  
- a **yellow dashed reference line** showing the world average depression level  
- a table that lists:
  - depression in 1990  
  - depression in 2009  
  - the percentage differences  

::: {.center}
<div style="text-align:center; margin-top: 1rem; margin-bottom: 1rem;">
  <embed src="img2.pdf" type="application/pdf" width="100%" height="900px" />
</div>
:::

*Figure 2: Depression percentages in America (IHME GBD) for 2009 compared with 1990.*

###  Key insights
::: {.callout-tip}
- Several age groups show increases or decreases between 1990 and 2009.  
- The “Standardised” row provides an overall summary measure of change.  
- The global average line allows quick comparison between America and global levels.  
:::

:::

---

# 3. Coping Methods Comparison — Between Countries

::: {.card style="padding: 1.2rem; margin-bottom: 1.5rem;"}

###  What this visualisation shows

This tab compares **coping strategies** used in two selected countries.

The screenshot below compares **Australia** and **India**.  
The chart illustrates coping strategies such as:

- talking to friends or family  
- lifestyle improvements  
- spending time in nature  
- seeking professional help  
- medication use  
- work or social adjustments  

::: {.center}
<div style="text-align:center; margin-top: 1rem; margin-bottom: 1rem;">
  <embed src="img3.pdf" type="application/pdf" width="100%" height="500px" />
</div>
:::



*Figure 3: Comparison of coping methods between Australia and India.*

###  Key insights
::: {.callout-tip}
- Australia shows greater use of professional services and lifestyle-based coping methods.  
- India shows stronger use of community and family-based support strategies.  
- The ranked bar chart clearly highlights cultural differences in coping preferences.  
:::

:::

---

# Full Vignette Documentation

For a complete walkthrough of the app, dataset, functions, and development process, refer to:

 **[Full App Overview Documentation](https://etc5523-2024.github.io/assignment-4-packages-and-shiny-apps-praj0022/)**

---

# Run the App Locally

```{r, eval=FALSE}
library(copingmethods)
launch_copingmethods()
```

Launching this command will open the interactive Shiny app in your browser.

---

# Why This Page Exists

Because hosting Shiny apps online requires paid servers,  
this documentation page ensures the app is still fully understandable by providing:

- static screenshots  
- clear explanations  
- highlighted insights  
- a faithful visual representation of the interface  

This allows assessors, collaborators, and website viewers to understand the app without needing to run it immediately.
