---
title: "From Isolation to Connection: How the World Managed Mental Health in 2020"
author: "Pooja Rajendran Raju"
date: "2024-10-16"
categories: [Blog]
---

```{r Setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(ggimage)
library(rsvg)
library(tidytext)
```

When COVID-19 hit, the world didn’t just face a health crisis—it also grappled with a mental health crisis. Depression rates surged as people dealt with isolation, financial worries, and uncertainty. So, how did people cope? That’s what we’re digging into here: the different ways people across the globe managed their mental health during the pandemic and what were the most popular coping stategies. By understanding these unique strategies, we can build better mental health support systems that fit different cultural and social contexts. 

Depression isn’t confined to any one group—it impacts people of all ages, backgrounds, and cultures. Recent data shows that depressive symptoms were already on the rise by 2019, particularly affecting older age groups. The COVID-19 pandemic only intensified this trend, with anxiety and depression surging worldwide as observed in @fig-trend. According to the World Health Organization (2022), mental health became a critical issue for millions, emphasizing the urgent need for support.

```{r}
# Loading data
data_1 <- read.csv("data/depressive-disorders-prevalence-by-age.csv")
```

```{r}

# Cleaning the data 
tidy_data_1 <- data_1 |> 
               rename( Age_5_14 = Depressive.disorders..share.of.population....Sex..Both...Age..5.14.years, 
                     Age_15_19 = Depressive.disorders..share.of.population....Sex..Both...Age..15.19.years, 
                     Age_20_24 = Depressive.disorders..share.of.population....Sex..Both...Age..20.24.years, 
                     Age_25_29 = Depressive.disorders..share.of.population....Sex..Both...Age..25.29.years, 
                     Age_30_34 = Depressive.disorders..share.of.population....Sex..Both...Age..30.34.years, 
                     Age_35_39 = Depressive.disorders..share.of.population....Sex..Both...Age..35.39.years, 
                     Age_40_44 = Depressive.disorders..share.of.population....Sex..Both...Age..40.44.years, 
                     Age_45_49 = Depressive.disorders..share.of.population....Sex..Both...Age..45.49.years, 
                     Age_50_54 = Depressive.disorders..share.of.population....Sex..Both...Age..50.54.years, 
                     Age_55_59 = Depressive.disorders..share.of.population....Sex..Both...Age..55.59.years, 
                     Age_60_64 = Depressive.disorders..share.of.population....Sex..Both...Age..60.64.years, 
                     Age_65_69 = Depressive.disorders..share.of.population....Sex..Both...Age..65.69.years, 
                     Age_70 = Depressive.disorders..share.of.population....Sex..Both...Age..70..years, 
                     All_Ages = Depressive.disorders..share.of.population....Sex..Both...Age..All.ages , 
                     Age_Standardised = Depressive.disorders..share.of.population....Sex..Both...Age..Age.standardized)
```

```{r}
#Calculating average share of various age groups with depression around the world in 2019 

avg_share <- tidy_data_1 |>
             filter( Year == 2019) |>
             summarise( Age_5_14 = mean(Age_5_14), 
                        Age_15_19 = mean(Age_15_19), 
                        Age_20_24 = mean(Age_20_24), 
                        Age_25_29 = mean(Age_25_29), 
                        Age_30_34 = mean(Age_30_34), 
                        Age_35_39 = mean(Age_35_39), 
                        Age_40_44 = mean(Age_40_44), 
                        Age_45_49 = mean(Age_45_49),
                        Age_50_54 = mean(Age_50_54), 
                        Age_55_59 = mean(Age_55_59), 
                        Age_60_64 = mean(Age_60_64), 
                        Age_65_69 = mean(Age_65_69), 
                        Age_70 = mean(Age_70), 
                        All_Ages = mean(All_Ages))

long_data <- avg_share |> 
  pivot_longer(
    cols = starts_with("Age_"),
    names_to = "age_groups",
    values_to = "share_avg"
    ) |> 
  select(-All_Ages)
                        
long_data <- long_data |>
             mutate(share_avg = round(share_avg, 1))                        
```


```{r fig.width= 11, fig.height = 6 }
#| label: fig-trend
#| fig-cap: "Prevalence of Depression by Age Group"
# Creating the plot 

ggplot(long_data, aes(x = share_avg, y = reorder(age_groups, share_avg), fill = share_avg)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(share_avg, "%")), hjust = -0.1) +
  scale_fill_gradient(high = "maroon", low = "yellow") +
  theme_minimal() +
  labs(title = "Prevalence of Depression by Age Group",
       x = "Percentage",
       y = "Age Group") +
  theme(axis.text.x = element_text(angle = 0, hjust = 1), 
        plot.margin = margin(t = 20, r = 40, b = 20, l = 20),
        legend.position = "none")


```

What’s fascinating is how people around the world responded to these mental health challenges. Cultural norms, societal values, and economic realities all influenced the ways people coped with anxiety and depression. Understanding these coping methods matters because it helps tailor mental health services to better address individual needs. It refines therapeutic approaches, making them more responsive to diverse preferences, and empowers people to build resilience by drawing on the coping practices of others. Additionally, these insights contribute to mental health research, paving the way for innovative therapies that make a meaningful difference.

It’s clear that a one-size-fits-all approach to mental health won’t work. The more we understand about different coping strategies, the better we can support those who need it most. To explore these coping strategies, I analyzed the 2020 dataset from Our [World in Data](https://ourworldindata.org/grapher/dealing-with-anxiety-depression-comparison?tab=table), “How do people deal with anxiety or depression around the world?” 

The dataset includes responses from people who experienced anxiety or depression severe enough to disrupt their daily lives for two weeks or more. Key variables include each country’s name (Entity), the country code (Code), and the year of reporting (Year). Most importantly, the dataset details the percentage of people engaging in various coping methods, like spiritual activities, healthy lifestyle changes, altering work or personal relationships, talking to loved ones, taking prescribed medication, spending time in nature, and seeking professional help.

To make sense of the data, I cleaned up the dataset by renaming columns to make them clearer and more descriptive. Next, I calculated the average percentage of use for each coping method to find out which strategies were the most common. I used the summarise function for this step and took care of any missing values along the way. Once I had the averages, I transformed the data into a long format to make it easier to work with. From there, I created a summary table to show the share of use for each coping strategy, formatting it with kable so it was neat and easy to read.

But I wanted to go beyond averages. To see how coping strategies varied across countries, I reshaped the data into a long format again, where each row represented a country and its corresponding share for each coping method. After grouping by method using group_by function and sorting in descending order, I identified the top 5 countries for each coping strategy using the top_n function.

I visualized these insights with ggplot, creating a bar plot with the percentage share of each method on the x-axis and the corresponding countries on the y-axis. I used facet_wrap to break down the plot by coping method, presenting methods into separate panels for clarity, with two columns.

**Top Three Coping Methods**

Coping with anxiety and depression varies widely, shaped by factors like environment, cultural background, and access to resources. Yet, despite the differences some strategies stand out as widely adopted across the globe. So, which coping methods were the most commonly used? Let’s explore the top three ways people around the world managed their mental health during the pandemic.

```{r}
# Reading the data
data_2 <- read.csv("data/dealing-with-anxiety-depression-comparison.csv")
```

```{r}
# Data Cleaning
tidy_data_2 <- data_2 |>
               rename(Religious_spiritual_activities =  Share...Question..mh8b...Engaged.in.religious.spiritual.activities.when.anxious.depressed...Answer..Yes...Gender..all...Age.group..all, 
              Improved_healthy_lifestyle_behaviors = Share...Question..mh8e...Improved.healthy.lifestyle.behaviors.when.anxious.depressed...Answer..Yes...Gender..all...Age.group..all, 
              Made_a_change_to_work_situation = Share...Question..mh8f...Made.a.change.to.work.situation.when.anxious.depressed...Answer..Yes...Gender..all...Age.group..all, 
              Made_a_change_to_personal_relationships = Share...Question..mh8g...Made.a.change.to.personal.relationships.when.anxious.depressed...Answer..Yes...Gender..all...Age.group..all, 
              Talked_to_friends_or_family = Share...Question..mh8c...Talked.to.friends.or.family.when.anxious.depressed...Answer..Yes...Gender..all...Age.group..all, 
              Took_prescribed_medication = Share...Question..mh8d...Took.prescribed.medication.when.anxious.depressed...Answer..Yes...Gender..all...Age.group..all, 
              Spent_time_in_nature = Share...Question..mh8h...Spent.time.in.nature.the.outdoors.when.anxious.depressed...Answer..Yes...Gender..all...Age.group..all, 
              Talked_to_mental_health_professional = Share...Question..mh8a...Talked.to.mental.health.professional.when.anxious.depressed...Answer..Yes...Gender..all...Age.group..all)
```

```{r}
#| label: tbl-coping
#| tbl-cap: "Share of Use of Various Coping Methods"
# Calculating the top 3 strategies used to cope with depression 

top_five <- tidy_data_2 |> 
  summarise(
    "Religious spiritual activities" = format(round(mean(Religious_spiritual_activities, na.rm = TRUE), 2), nsmall = 2), 
    "Improved healthy lifestyle behaviors" = format(round(mean(Improved_healthy_lifestyle_behaviors, na.rm = TRUE), 2), nsmall = 2), 
    "Made a change to work situation" = format(round(mean(Made_a_change_to_work_situation, na.rm = TRUE), 2), nsmall = 2), 
    "Made a change to personal relationships" = format(round(mean(Made_a_change_to_personal_relationships, na.rm = TRUE), 2), nsmall = 2), 
    "Talked to friends or family" = format(round(mean(Talked_to_friends_or_family, na.rm = TRUE), 2), nsmall = 2), 
    "Took prescribed medication" = format(round(mean(Took_prescribed_medication, na.rm = TRUE), 2), nsmall = 2),
    "Spent time in nature" = format(round(mean(Spent_time_in_nature, na.rm = TRUE), 2), nsmall = 2), 
    "Talked to mental health professional" = format(round(mean(Talked_to_mental_health_professional, na.rm = TRUE), 2), nsmall = 2)
  )


table_data <- top_five |>
  pivot_longer(
   cols = matches("^[A-Z]"),
    names_to = "Methods",
    values_to = "Percentage_Share"
    ) |>
  arrange(desc(Percentage_Share)) 
  

 table_data |>
        kable(col.names = c("Methods", "Share of Use in Percentage"), 
        caption = "Share of Use of Various Coping Methods",
          format = "html", align = c("l", "c")) |>
        kable_styling(bootstrap_options = c("striped"), 
                full_width = F) |>
        row_spec(1:3, bold = TRUE) 
        
``` 

Among the eight strategies analyzed in  @tbl-coping, talking to friends or family, spending time in nature, and improving healthy lifestyle behaviors took the top spots. Turns out, 79.5% of people leaned on their social circles, choosing to talk to friends or family when managing anxiety or depression. It makes sense—social support is a tried-and-true way to feel less isolated and more grounded when things get tough. Meanwhile, 72.01% of people found comfort in the great outdoors. Spending time in nature isn’t just a nice escape; it's scientifically proven to lower stress and boost your mood. Whether it’s a walk in the park or a weekend hike, the fresh air works wonders. Then there’s improving healthy lifestyle behaviors, chosen by 70.85%. Small changes like exercising more, eating well, and getting enough sleep can make a significant difference in how people manage their mental health, highlighting the important link between physical and emotional well-being.

**For Each Coping Strategy, What Were The Top 5 Countries They Were Most Popular In**

Different countries handle mental health in their own unique ways. Just take a look at @fig-country—the coping strategies vary a lot depending on where you are, and that tells us something important about exploring various strategies to customise interventions for diverse demographic, ethnic, and socio-economic contexts  (Bell, Williams, & Cooper, 2021). 
```{r}

ind_country <- tidy_data_2 |> 
           rename('Religious spiritual activities' = Religious_spiritual_activities, 
              'Improved healthy lifestyle behaviors' = Improved_healthy_lifestyle_behaviors , 
              'Made a change to work situation' = Made_a_change_to_work_situation, 
              'Made a change to personal relationships' = Made_a_change_to_personal_relationships, 
              'Talked to friends or family' = Talked_to_friends_or_family, 
              'Took prescribed medication' = Took_prescribed_medication, 
              'Spent time in nature' = Spent_time_in_nature, 
              'Talked to mental health professional' = Talked_to_mental_health_professional) |>
  pivot_longer(
   cols = starts_with("R") | starts_with("I") | starts_with("M") | starts_with("T") | starts_with("S"),
    names_to = "Methods",
    values_to = "Percentage_Share"
  ) 

ind_country <- ind_country |>
                mutate(Percentage_Share = round(Percentage_Share, 1))       

ind_country <- ind_country |> 
               group_by(Methods) |> 
               arrange(desc(Percentage_Share)) |> 
               top_n(5)
  

              
```


```{r}
ind_country <- ind_country |>
  mutate(Flag = case_when(
    Entity == "Thailand" ~ "https://flagcdn.com/w320/th.png",
    Entity == "Ecuador" ~ "https://flagcdn.com/w320/ec.png",
    Entity == "Laos" ~ "https://flagcdn.com/w320/la.png",
    Entity == "Nicaragua" ~ "https://flagcdn.com/w320/ni.png",
    Entity == "Peru" ~ "https://flagcdn.com/w320/pe.png",
    Entity == "El Salvador" ~ "https://flagcdn.com/w320/sv.png",
    Entity == "Indonesia" ~ "https://flagcdn.com/w320/id.png",
    Entity == "Nigeria" ~ "https://flagcdn.com/w320/ng.png",
    Entity == "Sri Lanka" ~ "https://flagcdn.com/w320/lk.png",
    Entity == "Kenya" ~ "https://flagcdn.com/w320/ke.png",
    Entity == "Mali" ~ "https://flagcdn.com/w320/ml.png",
    Entity == "Zambia" ~ "https://flagcdn.com/w320/zm.png",
    Entity == "Philippines" ~ "https://flagcdn.com/w320/ph.png",
    Entity == "Zimbabwe" ~ "https://flagcdn.com/w320/zw.png",
    Entity == "Sweden" ~ "https://flagcdn.com/w320/se.png",
    Entity == "United States" ~ "https://flagcdn.com/w320/us.png",
    Entity == "Albania" ~ "https://flagcdn.com/w320/al.png",
    Entity == "Italy" ~ "https://flagcdn.com/w320/it.png",
    Entity == "United Kingdom" ~ "https://flagcdn.com/w320/gb.png",
    Entity == "Finland" ~ "https://flagcdn.com/w320/fi.png",
    Entity == "Netherlands" ~ "https://flagcdn.com/w320/nl.png",
    Entity == "Switzerland" ~ "https://flagcdn.com/w320/ch.png",
    Entity == "Canada" ~ "https://flagcdn.com/w320/ca.png",
    Entity == "France" ~ "https://flagcdn.com/w320/fr.png",
    Entity == "Spain" ~ "https://flagcdn.com/w320/es.png",
    Entity == "Dominican Republic" ~ "https://flagcdn.com/w320/do.png",
    TRUE ~ NA_character_  # Catch-all for any unmatched countries
  ))
```

```{r fig.width= 120, fig.height = 70}
#| label: fig-country
#| fig-cap: "Top Preferences of Coping Methods by Country"

ggplot(ind_country, aes(x = Percentage_Share, y = reorder_within(Entity, Percentage_Share,Methods),  fill = Percentage_Share)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(Percentage_Share, "%")), hjust = -0.1, size = 25, fontface = "bold") +
  theme_minimal() +
  labs(title = "Strategies for Coping with Anxiety and Depression in 2020", 
       x = "Percentage", y = "Country") +
  facet_wrap(~ Methods, ncol = 2, scales = "free_y") +
  scale_fill_gradient(low = "lightblue", high = "darkblue")  +
  geom_image(aes(x = Percentage_Share - 5, image = Flag), size = 0.2)  + 
    scale_y_reordered() +
  theme(strip.text = element_text(size = 70, face = "bold"),
        plot.title = element_text(size = 90, hjust = 0.5, face = "bold"),
        axis.text.x = element_text(size = 60, face = "bold"),
        axis.ticks.x = element_blank(),
        axis.text.y = element_text(size = 70, face = "bold"),
        panel.spacing = unit(3, "lines"),
        axis.title.x = element_text(size = 90, face = "bold"),      
        axis.title.y = element_text(size = 90, face = "bold"),      
        legend.text = element_text(size = 75, face = "bold"),       
        legend.title = element_text(size = 90, face = "bold"),
        legend.key.size = unit(10, "lines"))

```

In places like Ecuador, Peru, and Nicaragua, people tend to focus on improving lifestyle habits, changing personal relationships, or tweaking their work situations. Basically, it’s all about making adjustments in their personal lives to manage mental health. Now, compare that to the United Kingdom, Switzerland, and Canada, where the approach leans more on professional support. Here, you see people taking prescribed medication or reaching out to mental health professionals. It makes sense, given the healthcare systems in these countries make it easier to access professional help.

In the United States, Italy, and Albania, spending time in nature is a popular way to cope with mental health challenges. It’s not just about getting outside—it’s recognizing the stress-relieving benefits of the natural world and how it helps boost mental well-being. Meanwhile, in places like Kenya, Nigeria, and Sri Lanka, religious or spiritual activities are a common go-to. It highlights how, in these lower-middle-income countries, mental health care is often intertwined with cultural and spiritual traditions. And then you’ve got the Philippines, Zimbabwe, and Laos, where people are more likely to talk to friends or family. This shows just how much social support and community connections matter when it comes to managing anxiety and depression.

So what’s the takeaway? While talking to friends is one of the most common ways to deal with mental health issues, the coping strategies people use differ a lot from country to country. It all comes down to things like background, ethnicity, and economic conditions. In the end, coping often means a mix of social interaction, time outdoors, and healthy habits—all of which contribute to mental resilience. The results also remind us that mental health support isn’t one-size-fits-all. In wealthier countries, people might lean more on mental health professionals and medication, while in lower-middle-income nations, religious or spiritual practices play a bigger role.

Understanding these differences is key to creating mental health strategies that actually work for people, no matter where they’re from. It’s a big step toward developing culturally relevant interventions that can help individuals across all walks of life.


**References**

1. World Health Organization. (2022, March 2). COVID-19 pandemic triggers 25% increase in prevalence of anxiety and depression worldwide. World Health Organization. https://www.who.int/news/item/02-03-2022-covid-19-pandemic-triggers-25-increase-in-prevalence-of-anxiety-and-depression-worldwide
2. Bell, C. A., Williams, J. L., & Cooper, S. E. (2021). Coping with the COVID-19 pandemic: A review of mental health strategies. Journal of Global Health, 11(3), 456-467. https://www.jogh.org/coping-with-the-covid-19-pandemic-a-review-of-mental-health-strategies



